Distances

### 2020-04-20 03_distances.Rmd
Clean up distances file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
library(tidyverse)
library(sf)
library(leaflet)
use_condaenv("ox", conda = "~/opt/anaconda3/condabin/conda", required = TRUE)
```

@@

```{python setup}
import osmnx as ox
import networkx as nx

import numpy as np
import pandas as pd
import time
```

@@

We use the `osmnx` library to get a highway network graph for New York City from 
OpenStreetMap. We get the walk network because most people in New York City will
not be driving to get to their park.

```{python graph}
graph = ox.graph_from_place('New York, New York, USA', network_type='walk')
# project to 18N, EPSG = 32618
graph_proj = ox.project_graph(graph, to_crs="+proj=utm +zone=17 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
graph_simple = nx.DiGraph(graph_proj)
```

@@

We get a list of population-weighted block group centroids from Census.

```{python blockgroups}
# get the coordinates from census
url = "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG36.txt"
blockgroups = pd.read_csv(url)

# pad the county and tract strings, and get the proper data types
blockgroups.STATEFP = blockgroups.STATEFP.apply(str)
blockgroups.BLKGRPCE = blockgroups.BLKGRPCE.apply(str)
blockgroups.COUNTYFP = blockgroups.COUNTYFP.apply('{0:0>3}'.format)
blockgroups.TRACTCE = blockgroups.TRACTCE.apply('{0:0>6}'.format)
blockgroups['GEOID'] = blockgroups[['STATEFP', 'COUNTYFP', 'TRACTCE', 'BLKGRPCE']].apply(lambda x: ''.join(x), axis = 1)

blockgroups['LATITUDE'] = blockgroups['LATITUDE'].astype(float)
blockgroups['LONGITUDE'] = blockgroups['LONGITUDE'].astype(float)

# filter to nyc
nyc_counties = ["081", "047", "061", "005", "085"]
blockgroups = blockgroups[blockgroups.COUNTYFP.isin(nyc_counties)]
blockgroups.set_index('GEOID', inplace=True)
```

@@

Get the data into a form that we can pass it into python.

@@

First, we bring the park points we sampled into a pandas data frame with
the park id as the index.
```{python park_points_py}
park_points = pd.DataFrame(r.park_points)
# TODO: run for all park points
park_ids = park_points.sample(5).id

park_points.set_index('id', inplace=True)
park_points = park_points.loc[park_ids]

```

We already created the block group centroids data frame earlier. Let's get the
nearest network nodes for each of the block groups and each park point.

```{python park_bg_nodes}
# TODO: run for all block groups
blockgroups = blockgroups.sample(5)

def findnode(df):
  df['node'] = df.apply(lambda row: 
    ox.get_nearest_node(graph, (row.LATITUDE, row.LONGITUDE)), axis=1)
  return df

a = time.process_time()
findnode(blockgroups)
findnode(park_points)
b = time.process_time()
b - a
```

We want to create a data structure that we can quickly populate with the 
shortest path between a block group and the nearest park point. This is 
probably a two-level dictionary.

```{python blockgroup_dict}
# Create a dictionary for the block groups
blockgroup_dict = {}
for i in blockgroups.index.unique():
  blockgroup_dict[i] = {} # create a dictionary for the parks
  for j in park_points.index.unique():
    blockgroup_dict[i][j] = float("inf")

```

Now we want to loop through each block group and all of the park points to figure
out what the shortest path is.

```{python nearest}
# loop through block groups
a = time.process_time()
for bg in blockgroups.itertuples():

  # loop through parks
  for park in park_ids:
    these_points = park_points.loc[[park]]

    # loop through points
    for point in these_points.itertuples():
      # get length between park and this edge point
      try:
        length = nx.shortest_path_length(graph_proj, source = bg.node, target=point.node,
          weight = 'length')
      except:
        length = float("inf")

      # check if this is the shortest one we've found; if so, update.
      if length < blockgroup_dict[bg.Index][park]:
        blockgroup_dict[bg.Index][park] = length

b = time.process_time()
b - a 
```

```{python write}
park_points.to_csv("data/park_points.csv")
blockgroups.to_csv("data/blockgroups.csv")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, python.reticulate = FALSE, cache = TRUE)
library(reticulate)
library(tidyverse)
library(sf)
library(leaflet)
use_condaenv("ox", conda = "~/opt/anaconda3/condabin/conda", required = TRUE)
```

@@

 > The python calculations in this page are illustrations of the methods
 performed; the code is not actually literate as the computations do not complete 
 in a reasonable amount of time. The actual python script is in [py/park_distances.py](py/park_distances.py)

```{python setuppy}
import osmnx as ox
import networkx as nx

import numpy as np
import pandas as pd
import time
import pickle
```

@@

We use the `osmnx` library to get a highway network graph for New York City from 
OpenStreetMap. We get the walk network because most people in New York City will
not be driving to get to their park. We then project the network into UTM 17N 
so that all paths will return in meters.

```{python graph}
def get_graph(place, mode, crs):
    """
    Get a graph for the place and write both the original graph and the projected directed
    graph to file
    :param place:
    :param mode: One of "drive", "walk", etc.
    :return: OSMNX projected graph
    """
    # Get the graph and make it a non multigraph
    graph = ox.graph_from_place(place, network_type=mode)
    # write to file
    with open("data/graph.file", "wb") as f:
        pickle.dump(graph, f)
    # project to UTM zone 18 N and simplify
    graph_proj = ox.project_graph(graph, to_crs=crs)
    graph_proj = nx.DiGraph(graph_proj)
    with open("data/graph_proj.file", "wb") as f:
        pickle.dump(graph_proj, f)
    return graph_proj
```

An image of this graph is shown below, zoomed in on Midtown Manhattan / Columbus
Circle.

```{r columbus-circle}
knitr::include_graphics("figs/nyc_network_columbus_circle.png")
```

@@

The tract and socioeconomic information is provided at the tract level, but we
can calculate the accessibility at the block group level. We get a list of
population-weighted block group centroids from Census, and transform them into 
UTM17N coordinates.

```{r blockgroups}
# get the coordinates from census
url = "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG36.txt"
bg <- read_csv(url, col_types = list(
  STATEFP = col_character(),
  COUNTYFP = col_character(),
  TRACTCE = col_character(),
  BLKGRPCE = col_character()
)) %>%
  filter(COUNTYFP %in% c("081", "047", "061", "005", "085")) %>%
  transmute(GEOID = str_c(STATEFP, COUNTYFP, TRACTCE, BLKGRPCE),
            LATITUDE, LONGITUDE) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326) %>%
  st_transform(32618) %>%
    mutate(
    LONGITUDE = st_coordinates(.)[, 1],
    LATITUDE = st_coordinates(.)[,2]
  ) %>%
  st_set_geometry(NULL) %>%
  as_tibble() %>%
  write_csv("data/blockgroups_nonode.csv")
```

@@

Now, we transform the points into UTM Zone 17N.

@@

In order to calculate the distances between the block group centroids and the 
park edge points, we need to look up the nearest point on the network to each
point.  First, we bring the park points we sampled into a pandas data frame with
the park id as the index.

```{python park_points_py}
def find_node(df):
    """
    Function to find the nearest node in a network to lat,long columns in data frame
    :param df:  Pandas data frame with lat, long information
    """
    with open("data/graph_proj.file", "rb") as f:
        graph = pickle.load(f)
    df['node'] = df.apply(lambda row:
      ox.get_nearest_node(graph, (row.LATITUDE, row.LONGITUDE)), axis=1)
    return df
```

Now we get the shortest paths. The network is quite large with 122,794 nodes and
381,656 edges. We also have `r nrow(bg)`  and `r nrow(park_points)`, meaning
we need to find `r nrow(bg) * nrow(park_points)` shortest paths. Even on my beefy
computer this can cause some problems. So we use the Euclidean distance if over 
3.5 km, and only get the shortest paths for points less than 3.5km away.

```{python nearest}
# loop through block groups
def get_shortest_paths(df):
    """
    :param df:
    """
    # get graph information
    print("Getting graph information")
    with open("data/graph_proj.file", "rb") as f:
        graph = pickle.load(f)
    # output file
    write_file = "data/shortest_paths_" + str(os.getpid()) + ".csv"
    f = open(write_file, "w+")
    f.write("geoid, park_id, distance\n")
    for bg in df.itertuples():
        # loop through parks
        for park in park_ids:
            these_points = park_points.loc[[park]]
            # loop through points
            min_dist = float("inf")
            min_euc = float("inf")
            for point in these_points.itertuples():
                # check euclidean distance
                dx = point.LONGITUDE - bg.LONGITUDE
                dy = point.LATITUDE - bg.LATITUDE
                euc_dist = math.sqrt(dx**2 + dy**2)
                # if the Euclidean distance is more than 3500 meters, we will just use that
                if euc_dist > 3500:
                    if euc_dist < min_euc:
                      min_euc = euc_dist
                    break
                # If it's within 3500m Euclidean, then we will get the network distance
                try:
                    length = nx.shortest_path_length(graph, source=bg.node, target=point.node, weight='length')
                except: # unless there is no path
                    length = float("inf")
                # check if this is the shortest one we've found; if so, update.
                if length < min_dist:
                    min_dist = length
            # write out the length between the block group and this park
            used_distance = min(min_dist, min_euc)
            f.write(str(bg.Index) + ", " + str(park) + ", " + str(used_distance) + "\n")
    # close the buffer
    f.close()
```
